---
title: Logistic regression with multiple predictors
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Scenario: dengue fever in Vietnam

.large[
* *Dengue fever*: a mosquito-borne viral disease, which infects hundreds of millions of people a year. Common in tropical climates
* Researchers in Vietnam are interested in identifying the relationship between specific symptoms and the probability of having dengue
* Lab tests are available, but may take time to return results and/or be expensive
]

---

### Data

.large[
Data on 5720 Vietnamese children, admitted to hospital with possible dengue fever. Variables include:
* `Dengue`: whether the patient actually has dengue fever, based on a lab test (0 = no, 1 = yes)
* `Temperature`: patient's body temperature (in Celsius)
* `Abdominal`: whether the patient has abdominal pain (0 = no, 1 = yes)
* `HCT`: patient's hematocrit (proportion of red blood cells)
* `Age`: patient's age (in years)
* `Sex`: patient's sex 
* `+` several others
]

---

### Research question

.large[
You are approached by the researchers to help analyze their data. Their initial question: 

*What is the relationship between temperature and the probability of dengue, and does the relationship differ if the patient also presents with abdominal pain?*
]

---

### Exploratory data analysis

.large[
.question[
How can we visualize the relationship between temperature and the probability of having dengue?
]
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=5, fig.align='center'}
library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       #x <- c(head(x, lines), more)
       x <- c(head(x, lines))
     }
   } else {
     #x <- c(more, x[lines], more)
     x <- c(x[lines])
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")

logodds_plot <- function(data, num_bins, bin_method,
                         x_name, y_name, grouping = NULL, reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = 1)
  } else {
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = as.factor(data %>% pull(grouping)))
  }
  
  if(bin_method == "equal_size"){
    logodds_table <- dat %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  } else {
    logodds_table <- dat %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  }
  
  if(is.null(grouping)){
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds") +
      theme(text = element_text(size=25))
  } else {
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds,
                 color = group,
                 shape = group)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds",
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=25))
  }
  
}


logodds_plot(dengue, 12, "equal_size", x = "Temperature", y = "Dengue")
```

.large[
.question[
Based on the empirical log odds plot, what would be a reasonable model for the relationship between temperature and the probability a patient has dengue?
]
]

---

### Initial model

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Temperature_i$$
]

.large[
We're also interested in whether this relationship is different depending on abdominal pain.
.question[
What plot can I use to investigate this question?
]
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "Temperature", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
Based on this plot, what would be a reasonable model that incorporates both temperature and abdominal pain?
]
]

---

### Adding abdominal pain

.large[
$$Y_i \sim Bernoulli(\pi_i)$$
]

.large[
\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ & \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
* The interaction term allows the relationship between temperature and $\pi_i$ to change depending on abdominal pain
]

---

### Class activity

.large[
[https://sta214-f22.github.io/class_activities/ca_lecture_10.html](https://sta214-f22.github.io/class_activities/ca_lecture_10.html)
]

---

### Class activity

.large[
\begin{align}
\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) &= -7.745 + 0.178 \ Temperature_i \\
& \hspace{0.5cm} - 6.129 \ Abdominal_i \\ 
& \hspace{0.5cm} + 0.166 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
.question[
What is the estimated probability of dengue for a patient with a temperature of 38C and abdominal pain?
]
]

---

### Class activity

.large[
\begin{align}
\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) &= -7.745 + 0.178 \ Temperature_i \\
& \hspace{0.5cm} - 6.129 \ Abdominal_i \\ 
& \hspace{0.5cm} + 0.166 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
.question[
For patients with abdominal pain, what is the estimated change in odds associated with an increase in temperature of 1C?
]
]

---

### Class activity

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
I want to test whether there is a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue.
.question[
What are my null and alternative hypotheses?
]
]

---

### Hypothesis testing

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
**Hypotheses:**

$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

.large[
.question[
Which type of test can I use to test these hypotheses ( Wald test, likelihood ratio test, or either)?
]
]

---

### Likelihood ratio test

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
**Hypotheses:**

$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

.large[
.question[
What are my full and reduced models?
]
]

---

### Full and reduced models

.large[
**Full model:**

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}

**Reduced model:**

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Temperature_i$$
]

---

### Test statistic

.large[
$$G = \text{deviance of reduced model } - \text{deviance of full model}$$
]

.large[
**Full model:**

```{r echo=F, output.lines = 12:13}
glm(Dengue ~ Temperature*Abdominal, family=binomial, data=dengue)
```
]

.large[
**Reduced model:**

```{r echo=F, output.lines = 9:10}
glm(Dengue ~ Temperature, family=binomial, data=dengue)
```
]

.large[
$$G =$$
.question[
How do I calculate a p-value for this test statistic?
]
]

---

### p-value

.large[
$$G = 6927 - 6914 = 13 \hspace{1cm} G \sim \chi^2_k$$
where $k =$ difference in number of parameters between full and reduced models.

.question[
What is $k$ for this test?
]
]

---

### p-value

.large[
$$G = 6927 - 6914 = 13 \hspace{1cm} G \sim \chi^2_k$$
where $k =$ difference in number of parameters between full and reduced models.

```{r}
pchisq(13, df=2, lower.tail=F)
```
]

---

### Conclusion

.large[
* **Question:** Is there a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue?
* **Hypotheses:**
$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
* **p-value:** 0.0015
* **Conclusion:** There is strong evidence that there is a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue.
]

---

### A new question...

.large[
You report your results to the hospital, and they ask a follow-up question: 

*Does the model improve when we add hematocrit (the proportion of red blood cells)?*
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
Does it look like we need an interaction between hematocrit (HCT) and abdominal pain?
]
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
How can I check whether there might be an interaction between temperature and hematocrit?
]
]

---

### Exploratory data analysis

.large[
Define $TempHigh$ by 
* $TempHigh_i = 1$ if $Temperature_i > 38$
* $TempHigh_i = 0$ if $Temperature_i <= 38$
]

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
dengue %>%
  mutate(TempHigh = as.factor((Temperature > 38) + 0)) %>%
  logodds_plot(10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "TempHigh")
```

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
dengue %>%
  mutate(TempHigh = as.factor((Temperature > 38) + 0)) %>%
  logodds_plot(10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "TempHigh")
```

.large[
.question[
Does it look like we need an interaction between temperature and hematocrit?
]
]

---

### Model

.large[
$$Y_i \sim Bernoulli(\pi_i)$$
]

.large[
\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i \\
& \hspace{0.5cm} + \beta_4 \ HCT_i
\end{align}
]

.large[
* Note that while we binarized temperature for EDA, we use the original variable in the model here
]
