---
title: Logistic regression interpretation
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=F}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

### Class activity, Part I

.large[
[https://sta214-f22.github.io/class_activities/ca_lecture4.html](https://sta214-f22.github.io/class_activities/ca_lecture4.html)
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.901 + 0.0036 \ Score_i$$

.question[
Calculate the odds ratio comparing odds of acceptance for a student with a GRE score of 701 to a student with a GRE score of 700.
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.901 + 0.0036 \ Score_i$$

.question[
Calculate the odds ratio comparing odds of acceptance for a student with a GRE score of 701 to a student with a GRE score of 700.
]
]

.large[
$$\dfrac{e^{-2.901 + 0.0036 \cdot 701}}{e^{-2.901 + 0.0036 \cdot 700}} = 1.0036$$

.question[
This is the same odds ratio as comparing a 601 GRE to a 600 GRE!
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.901 + 0.0036 \ Score_i$$

Odds ratio for increasing GRE by one point:

]

---

### Interpreting coefficients

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$$

Interpretation of $\widehat{\beta}_1$ :

* A one unit increase in $x$ is associated with an increase of $\widehat{\beta}_1$ in the log odds
* A one unit increase in $x$ is associated with an increase in the odds by a factor of $\exp\{\widehat{\beta}_1\}$

.question[
How do you think we interpret $\widehat{\beta}_0$?
]
]

---

### Interpreting coefficients

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$$

Interpretation of $\widehat{\beta}_1$ :

* A one unit increase in $x$ is associated with an increase of $\widehat{\beta}_1$ in the log odds
* A one unit increase in $x$ is associated with an increase in the odds by a factor of $\exp\{\widehat{\beta}_1\}$

Interpretation of $\widehat{\beta}_0$ :

* The estimated log odds when $x = 0$ are $\widehat{\beta}_0$
* The estimated odds when $x = 0$ are $\exp\{\widehat{\beta}_0\}$
]

---

### Fitting logistic regression in R

.large[
```{r output.lines = 10:22}
gre_glm <- glm(admit ~ gre, data = grad_app,
               family = binomial)
summary(gre_glm)
```
]

---

### Class activity, Part II

.large[
[https://sta214-f22.github.io/class_activities/ca_lecture4.html](https://sta214-f22.github.io/class_activities/ca_lecture4.html)
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.358 + 1.051 \ GPA_i$$

.question[
What is the change in the odds of acceptance associated with an increase of 1 point in GPA?
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.358 + 1.051 \ GPA_i$$

.question[
What is the change in the odds of acceptance associated with an increase of 1 point in GPA?
]
]

.large[
An increase of 1 point in GPA is associated with an increase in the odds of acceptance by a factor of $\exp\{1.051\} = 2.861$
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.358 + 1.051 \ GPA_i$$

.question[
What is the estimated probability that a student with a GPA of 3.5 is accepted?
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.358 + 1.051 \ GPA_i$$

.question[
What is the estimated probability that a student with a GPA of 3.5 is accepted?
]
]

.large[
$\dfrac{e^{-4.358 + 1.051(3.5)}}{1 + e^{-4.358 + 1.051(3.5)}} \approx 0.336$
]

---

### Class activity

.large[
.question[
The logistic regression model assumes that the change in odds associated with an increase of 1 point in GPA is constant. Is the change in *probability* also constant?
]
]

---

### Class activity

.large[
.question[
The logistic regression model assumes that the change in odds associated with an increase of 1 point in GPA is constant. Is the change in *probability* also constant?
]
]

.large[
No:

* GPA = 2.0, estimated probability = 0.095
* GPA = 3.0, estimated probability = 0.231
* GPA = 4.0, estimated probability = 0.462
]