---
title: Zero inflated models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Recap: College drinking

.large[
Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: the number of drinks the student reports consuming
* `sex`: an indicator for whether the student identifies as male
* `OffCampus`: an indicator for whether the student lives off campus
* `FirstYear`: an indicator for whether the student is a first-year student

.question[
Our goal: model the number of drinks students report consuming.
]
]

---

### Recap: EDA

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)

wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
What do you notice about this distribution?
]
]

---

### Excess zeros

.large[
.question[
Why might a student report consuming 0 drinks?
]
]

---

### Zero-inflated Poisson (ZIP) model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

where

$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 FirstYear_i$

$\log(\lambda_i) = \beta_0 + \beta_1 OffCampus_i + \beta_2 Male_i$

.question[
What do $\alpha_i$ and $\lambda_i$ represent in this model?
]
]

---

### Fitting the model in R

.large[
```{r, output.lines = 10:20, message=F, warning=F}
library(pscl)
m1 <- zeroinfl(drinks ~ OffCampus + sex | FirstYear, 
               data = wdrinks)
summary(m1)
```
]

---

### Fitted ZIP model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -0.60 + 1.14 FirstYear_i$

$\log(\widehat{\lambda}_i) = 0.75 + 0.42 \ OffCampus_i + 1.02 \ Male_i$

.question[
How would I interpret the coefficient 1.14 in the fitted model?
]
]

---

### Fitted ZIP model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -0.60 + 1.14 FirstYear_i$

$\log(\widehat{\lambda}_i) = 0.75 + 0.42 \ OffCampus_i + 1.02 \ Male_i$

.question[
How would I interpret the coefficient 0.42 in the fitted model?
]
]

---

### Data: Framingham heart study

.large[
Data collected on residents of Framingham, MA over a long period of time, to study variables related to heart health. We will work with a subset of the data, containing

* `cigsPerDay`: The number of cigarettes smoked per day during the study period.
* `education`: 1 = High School, 2 = Some College, 3 = College Degree, 4 = Advanced Degree.
* `male`: 1 = Male, 0 = Female.
* `age`: The age of the individual in years.
* `diabetes`: 1 if the individual has diabetes, 0 otherwise.

]

---

### EDA: number of cigarettes smoked

```{r, echo=F, warning=F, message=F, fig.align='center', fig.height=5, fig.width=7}
heart_data <- read.csv("~/Documents/Teaching/sta214-f22.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay) %>%
  mutate(education = as.factor(education))

heart_data %>%
  ggplot(aes(x = cigsPerDay)) +
  geom_histogram(bins = 20) +
  theme_bw() +
  labs(x = "Number of cigarettes smoked per day") +
  theme(text = element_text(size = 20))
```

.large[
.question[
What do you notice about this distribution?
]
]

---

### EDA: number of cigarettes smoked

```{r, echo=F, warning=F, message=F, fig.align='center', fig.height=5, fig.width=7}

heart_data %>%
  ggplot(aes(x = cigsPerDay)) +
  geom_histogram(bins = 20) +
  theme_bw() +
  labs(x = "Number of cigarettes smoked per day") +
  theme(text = element_text(size = 20))
```

.large[
.question[
What latent (unobserved) binary variable would impact the number of cigarettes smoked?
]
]

---

### Class activity

.large[
[https://sta214-f22.github.io/class_activities/ca_lecture_23.html](https://sta214-f22.github.io/class_activities/ca_lecture_23.html)
]

---

### Class activity

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 2.93 -0.022 EducationSome_i -0.067 EducationCollege_i + \\ 0.009 EducationAdv_i - 0.046 Diabetes_i$$

.question[
How do we interpret the coefficient -0.046 in the fitted model?
]
]

---

### Class activity

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 2.93 -0.022 EducationSome_i -0.067 EducationCollege_i + \\ 0.009 EducationAdv_i - 0.046 Diabetes_i$$

.question[
What is the estimated probability that a 50 year old does not smoke?
]
]

---

### Class activity

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 2.93 -0.022 EducationSome_i -0.067 EducationCollege_i + \\ 0.009 EducationAdv_i - 0.046 Diabetes_i$$

.question[
What is the expected number of cigarettes smoked per day, for a smoker with diabetes and some college education?
]
]

---

### Class activity

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 2.93 -0.022 EducationSome_i -0.067 EducationCollege_i + \\ 0.009 EducationAdv_i - 0.046 Diabetes_i$$

.question[
What is the probability that a 45 year old college graduate without diabetes smokes one cigarette per day?
]
]

---

### Making predictions

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 2.93 -0.022 EducationSome_i -0.067 EducationCollege_i + \\ 0.009 EducationAdv_i - 0.046 Diabetes_i$$

.question[
How would I estimate the expected number of cigarettes smoked per day, by a college graduate without diabetes?
]
]

---

### A new question

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i$$

.question[
New research question: for smokers, does the number of cigarettes smoked per day depend on age?

How would we answer this research question?
]
]

---

### Inference

.large[
$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$

.question[
Research question: for smokers, does the number of cigarettes smoked per day depend on age?

What are the null and alternative hypotheses?
]
]

---

### Wald test

.large[
```{r, output.lines = 10:16}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
summary(m2)
```
]

---

### Likelihood ratio test

.large[

```{r}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
m2$loglik
```

```{r}
m1 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes | age, 
               data = heart_data)
m1$loglik
```
]