---
title: Prediction and hypothesis testing
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Data

.large[
**Question:** What is the relationship between age and contraceptive use for women in Indonesia?

**Data:** 1473 Indonesian couples, with variables
* $Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)
* $X_i$ = Wife's age (numeric)
]

---

### Last time: Fitted model

.large[
$$\log \left( \dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} \right) = -8.234 + 0.456 Age_i - 0.0065 Age_i^2$$

$$\log \left( \dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} \right) = -5.083 + 0.366 Age_i - 0.00628 Age_i^2$$
]

---

### Last time: Predictions

.large[
```{r include=F}
library(tidyverse)
library(nnet)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

cmc_data <- read_csv("../slides/cmc.data.txt", col_names = F)
colnames(cmc_data) <-c("WifeAge", "WifeEdu", "HusbandEdu", "NumChild", "WifeRel", "WifeWork", "HusbandOcc","Living", "Media", "Choice")
cmc_data$Choice <- ifelse( cmc_data$Choice == 1, "None", ifelse(cmc_data$Choice==2, "Short", "Long"))
cmc_data <- cmc_data %>%
  mutate(Choice = factor(Choice, levels=c("None", "Short", "Long")))

m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```

```{r echo=F}
preds <- predict(m1)
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```

.question[
How good are our predictions?
]

]

---

### Last time: Predictions

.large[

```{r echo=F}
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```

We can also assess our predictions by comparing to random guessing.

.question[
What are our predicted probabilities for each observation from random guessing?
]

]

---

### Random guessing

.large[
* If we don't have any data, our estimated probability would be 1/3 for each level
* If we have data but we don't use age, our estimated probability for each level is just the proportion of observations in that group:

```{r}
table(cmc_data$Choice)/nrow(cmc_data)
```
]

---

### Class activity

.large[
[https://sta214-f22.github.io/class_activities/ca_lecture_16.html](https://sta214-f22.github.io/class_activities/ca_lecture_16.html)
]

---

### Class activity

.large[
.question[
What would our confusion matrix look like if our predictions randomly assigned each person to one of the three categories, with a 1/3 chance for each category?
]
]

---

### Class activity

.large[
.question[
What would our confusion matrix look like if our predictions randomly assigned each person to one of the three categories, with a 1/3 chance for each category?
]
]

.large[
Something like 

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 210 | 111 | 170 |
| | Short | 210 | 111 | 170 |
| | Long | 209 | 111 | 171 |

]

---

### Class activity

.large[

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 210 | 111 | 170 |
| | Short | 210 | 111 | 170 |
| | Long | 209 | 111 | 171 |

.question[
What is the accuracy of our predictions in this confusion matrix?
]
]


---

### Class activity

.large[
.question[
What would our confusion matrix look like if for every individual, we just predicted the most common contraception choice in the data?
]
]

---

### Class activity

.large[
.question[
What would our confusion matrix look like if for every individual, we just predicted the most common contraception choice in the data?
]
]

.large[
The most common choice is None, so

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 629 | 333 | 511 |
| | Short | 0 | 0 | 0 |
| | Long | 0 | 0 | 0 |

]

---

### Class activity

.large[

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 629 | 333 | 511 |
| | Short | 0 | 0 | 0 |
| | Long | 0 | 0 | 0 |

.question[
What is the accuracy of our predictions in this confusion matrix?
]
]

---

### Class activity

.large[
.question[
Do we do better than random guessing?
]
]

---

### Moral

.large[
* By itself, accuracy isn't particularly useful for summarizing prediction performance
* It is helpful to interpret accuracy in relation to simple random guessing. Our model isn't very good if we can't beat a random guess
* We also need to look at predictive ability for each class
]

---

### Hypothesis testing

.large[
**Research question:** Is there a relationship between age and contraceptive choice?

.question[
What are my steps to answer this question with a hypothesis test?
]
]

---

### Specify hypotheses

.large[
**Research question:** Is there a relationship between age and contraceptive choice?

$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i + \beta_{2(Long)} Age_i^2$$

$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i + \beta_{2(Short)} Age_i^2$$

.question[
What should our null and alternative hypotheses be?
]

]

---

### Specify hypotheses

.large[
$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i + \beta_{2(Long)} Age_i^2$$

$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i + \beta_{2(Short)} Age_i^2$$

$H_0: \beta_{1(Short)} = \beta_{2(Short)} = \beta_{1(Long)} = \beta_{2(Long)} = 0$

$H_A: \text{at least one of } \beta_{1(Short)}, \beta_{2(Short)}, \beta_{1(Long)}, \beta_{2(Long)} \neq 0$

.question[
What are the full and reduced models?
]
]

---

### Test statistic

.large[
.question[
What test can I use to compare nested models?
]
]

---

### Drop in deviance

.large[
```{r include=F}
m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```

```{r eval=F}
m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```

```{r output.lines = 13:14}
summary(m1)
```

Deviance for full model: 3015.821

.question[
How would we fit the reduced model in R?
]
]
---

### Drop in deviance

.large[
```{r eval=F}
m0 <- multinom(Choice ~ 1, 
               data = cmc_data)
```

```{r include=F}
m0 <- multinom(Choice ~ 1, 
               data = cmc_data)
```

```{r output.lines = 13:14}
summary(m0)
```

Reduced model deviance: 3142.726

Drop in deviance: $G = 3142.726 - 3015.821 = 126.905$

.question[
What distribution do we use to calculate the p-value?
]
]

---

### Calculating a p-value

.large[

Under $H_0$, $G \sim \chi^2_q$, where $q$ is the number of parameters tested.

Here $q = 4$ (2 parameters for each log relative risk model)

```{r}
pchisq(126.905, df=4, lower.tail=F)
```

So we have very strong evidence that there is a relationship between age and contraceptive choice.
]

