---
title: Zero inflated models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(knitr)
library(pscl)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

heart_data <- read.csv("~/Documents/Teaching/sta214-f22.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay) %>%
  mutate(education = as.factor(education))
```


### A new question

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i$$

.question[
New research question: for smokers, does the number of cigarettes smoked per day depend on age?

How would we answer this research question?
]
]

---

### Inference

.large[
$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$

.question[
Research question: for smokers, does the number of cigarettes smoked per day depend on age?

What are the null and alternative hypotheses?
]
]

---

### Wald test

.large[
```{r, output.lines = 10:16}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
summary(m2)
```
]

---

### Likelihood ratio test

.large[

```{r}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
m2$loglik
```

```{r}
m1 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes | age, 
               data = heart_data)
m1$loglik
```
]