---
title: "Lab 8 Solutions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
bryozoan <- read_csv("https://sta112-s22.github.io/homework/bryozoan_data.csv")
```

### Question 1 (3 pts)

```{r}
bugula_early <- bryozoan %>%
  filter(Stage == "early", Species == "bugula")
```

**Grading:** I think many students might not filter by both species and stage, which will affect their results later in the lab.

* Lose 1 pt if they don't filter by species
* Lose 1 pt if they don't filter by stage
* They don't have to use the filter() function (e.g., subset would work too)


### Question 2 (2 pts)

```{r}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot() +
  labs(x = "Run", y = "Metabolic rate (mJ/hr)") +
  theme_bw()
```

There does seem to be a little variation between runs.

**Grading:** 1 pt for plot, 1 pt for interpretation


### Question 3 (2 pts)

There are several reasons why random effects are useful here:

* We're not interested in doing inference on the runs (that is, we don't care about comparing run 1 to run 2)
* There are 12 different runs, which would add 11 parameters to the model if we used fixed effects instead. We can get a simpler model, which allows us to interpret the variability between runs, by using a random effect for run
* We can imagine differences between runs are due to random variation in the experimental conditions, so a random effect makes intuitive sense. And with 12 runs, we probably have enough to estimate the random effect

**Grading:** Any of these explanations is acceptable


### Question 4 (3 pts)

```{r}
m1 <- lmer(Metabolic ~ (1|Run), data = bugula_early)
summary(m1)
```

$\widehat{\beta}_0 = 0.175$, $\widehat{\sigma}_u^2 = 0.00013$, $\widehat{\sigma}_\varepsilon^2 = 0.0042$

**Grading:** If they used the wrong subset in question 1, they will get different results


### Question 5 (2 pts)

The average metabolic rate, across runs, is 0.175 mJ/hr

### Question 6 (3 pts)

$\widehat{\rho}_{group} = \dfrac{0.00013}{0.00013 + 0.0042} = 0.03$

So, about 3% of variability in metabolic rate is explained by random variation between runs. This is very low, and smaller intra class correlations suggest a random effect is not as important.

**Grading:** 1 pt for correlation, 1 pt for interpretation, 1 pt for whether large or small correlation means random effect is important


### Question 7 (3 pts)

$$Metabolic_{ij} = \beta_0 + \beta_1 Mass_{ij} + u_i + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

**Grading:**

* Lose 1 pt for incorrect notation (e.g., missing subscripts)
* Lose 1 pt if they don't include the distribution of random effects and noise


### Question 8 (3 pts)

```{r}
m1 <- lmer(Metabolic ~ Mass + (1|Run), data = bugula_early)
summary(m1)
```

$\widehat{\beta}_0 = 0.071$, $\widehat{\beta}_1 = 0.0073$, $\widehat{\sigma}_u^2 = 0.00028$, $\widehat{\sigma}_\varepsilon^2 = 0.0036$


### Question 9 (3 pts)

The average metabolic rate when mass is 0 (i.e. the average intercept), across runs, is 0.071 mJ/hr. For any given run, an increase of 1 $\mu$g in mass is associated with an increase of 0.0073 mJ/hr in metabolic rate.


### Question 10 (3 pts)

```{r}
coef(m1)$Run[3,]
```

For Run 3, the estimated line is $0.053 + 0.0073 Mass_{ij}$. So, the predicted metabolic rate for a mass of 15 $\mu$g is $0.053 + 0.0073(15) = 0.163$ mJ/hr

### Question 11 (3 pts)

For a new run, the estimated line is $0.071 + 0.0073 Mass_{ij}$. So, the predicted metabolic rate for a mass of 15 $\mu$g is $0.071 + 0.0073(15) = 0.181$ mJ/hr.



