---
title: "Class Activity, September 26"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

## Setup

We'll continue analyzing the Indonesian contraception choice data that we were using in the slides.

# Part I

Our fitted multinomial regression model is

$$\log \left( \dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} \right) = -8.234 + 0.456 Age_i - 0.0065 Age_i^2$$

$$\log \left( \dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} \right) = -5.083 + 0.366 Age_i - 0.00628 Age_i^2$$

1. What is the predicted relative risk of short term use vs. no use for a woman aged 25?

2. What is the predicted *probability* of each contraceptive choice (None, Short, and Long) for a woman aged 25?

# Part II

Here is the confusion matrix for predictions from our fitted model:

```{r include=F}
library(tidyverse)
library(nnet)
library(knitr)

cmc_data <- read_csv("../slides/cmc.data.txt", col_names = F)
colnames(cmc_data) <-c("WifeAge", "WifeEdu", "HusbandEdu", "NumChild", "WifeRel", "WifeWork", "HusbandOcc","Living", "Media", "Choice")
cmc_data$Choice <- ifelse( cmc_data$Choice == 1, "None", ifelse(cmc_data$Choice==2, "Short", "Long"))
cmc_data <- cmc_data %>%
  mutate(Choice = factor(Choice, levels=c("None", "Short", "Long")))

m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```

```{r echo=F}
preds <- predict(m1)
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```

3. What fraction of our predictions are correct? (i.e., what is the *accuracy* of our predictions)

In the slides, we compared our predicted probabilities to random guessing. There were two ways we might randomly guess: without the data (assign a 1/3 chance of each contraception choice to every individual), or with the data (the probability of each choice is just the prevalence of that choice in the data).

4. What would our confusion matrix look like if our predictions randomly assigned each person to one of the three categories, with a 1/3 chance for each category?

5. What is the accuracy of our predictions in the confusion matrix for exercise 4?

6. What would our confusion matrix look like if for every individual, we just predicted the most common contraception choice in the data?

7. What is the accuracy of our predictions in the confusion matrix for exercise 6?

8. Does our model (with Age as a predictor) do better than random guessing?